---
title: "DATA 606 Project Proposal"
author: "Kristin Lussi"
date: "2023-10-29"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Data Preparation

```{r setup, echo=TRUE, results='hide', warning=FALSE, message=FALSE}
# load data
source("functions.r") # attached with submission
library(dplyr)
library(psych)
options(scipen=999)
wide_stack <- get_stack_df(persist = TRUE, load_from_cache = TRUE)
# specify the language columns
language_cols <- c("python", "sql", "java", "scala", "r", "julia")

wide_stack <- wide_stack %>%
  mutate(Gender = ifelse(Gender == "Woman", "Female", ifelse(Gender == "Man", "Male", Gender))) %>%
  filter(Year %in% 2018:2022) %>%
  filter(!is.na(AnnualSalary)) 
```

### Research question 

**You should phrase your research question in a way that matches up with the scope of inference your dataset allows for.**

What is the extent of the gender pay gap in the technology industry, and how does it vary across different ages, organization sizes, and regions?

### Cases 

**What are the cases, and how many are there?**

There are 44 cases in this data set, being: year, organization size (OrgSize), country, employment, gender, education level (EdLevel), Age, developer type (DevType), database knowledge (DatabaseWorkedWith), language knowledge (LanguageWorkedWith), platform knowledge (PlatformWorkedWith), years coding professionally (YearsCodePro), Annual Salary, YearsCodeProAvg, OrgSizeAvg, AgeAvg, etc. The remaining cases are columns that specify whether or not a person knows a language, database, or platform. There are 460,298 observations in this dataset. 


### Data collection 

**Describe the method of data collection.**

The data collection method for the Stack Overflow survey primarily involves conducting an annual online survey. Participants are invited to complete a web-based questionnaire, which is distributed to users of the Stack Overflow platform. This survey covers various aspects of software development, including job roles, programming languages, tools, technologies, and salary information. Participants voluntarily share their insights and experiences by responding to the questionnaire.

### Type of study 

**What type of study is this (observational/experiment)?**

This is an observational study. 

### Data Source 

**If you collected the data, state self-collected. If not, provide a citation/link.**

Data is collected by Stack Overflow and is available online here: https://insights.stackoverflow.com/survey


### Dependent Variable

**What is the response variable? Is it quantitative or qualitative?**

The response variable would be annual salary (quantitative).

### Independent Variable(s)

The independent variables that I will be analyzing in this project are:

- Gender (qualitative)
- Organization Size (quantitative)
- Country (qualitative)
- Age (quantitative)

### Relevant summary statistics 

**Provide summary statistics for each the variables. Also include appropriate visualizations related to your research question (e.g. scatter plot, boxplots, etc). This step requires the use of R, hence a code chunk is provided below. Insert more code chunks as needed.**

```{r split-genders}
# Split the data into two subsets by gender
males <- subset(wide_stack, Gender == "Male")
females <- subset(wide_stack, Gender == "Female")
```

```{r male-salary, warning = FALSE, message = FALSE}
describe(males$AnnualSalary)
```

```{r female-salary, warning = FALSE, message = FALSE}
describe(females$AnnualSalary)
```

```{r males-org size, warning = FALSE, message = FALSE}
describe(males$OrgSizeAvg)
```

