---
title: A Comprehensive Analysis of Gender Pay Gap in the Tech Industry
author: "Kristin Lussi and Tony Fraser"
date: "December 7th, 2023"
output:
  slidy_presentation: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Abstract
<!-- Summarize analysis -->
<!-- Research methods -->
<!-- Findings/conclusion -->
<!-- Significance of our findings -->

* Summary

This research paper uses six years' worth of the annual Stack Overflow survey data to examine the gender pay gap within the technology sector. It's important to note that the self-reported responses in this dataset are not necessarily specific to any particular industry but rather focus on individuals' job functions. This data spans a variety of tech positions such as R programmer, cloud engineer, technical project manager, blockchain developer, etc.

* Research Methods

In this research paper, you will find an overview of descriptive statistics describing the data and specific examples of the gender pay gap. Additionally, we conduct a deep dive using simple and multiple linear regression for some of our more advanced analyses.

* Findings and Conclusion

The data strongly indicates the presence of a substantial pay gap. However, despite the extensive nature of this dataset, it is apparent that there is at least one crucial yet unidentified variable that needs to be incorporated to effectively model the pay gap. We propose that without the inclusion of the "CompanyPercentSexist" column in this dataset, the comprehensive understanding and modeling of this pay gap may remain challenging.

* Significance of Findings

At the very least, we need to delve further into this issue. We would like to attempt to collect some funding and perform additional modeling. Our initial ideas involve presenting some additional questions to Stack Overflow, particularly related to geography and industry. While we might not be able to find "CompanyPercentSexist," narrowing our focus to industry and region could enable us to provide essential information to local politicians and the media. 

<!-- 
By using a data set that includes responses from a range of tech professionals, our goal is to offer a detailed insight into the factors that contribute to the identified trends in the gender pay gap. 
The results of this study could guide evidence-based efforts to promote inclusivity and equality, ultimately benefiting both the technology sector and society as a whole. -->

## Introduction

The gender pay gap within the US tech sector has long been a subject of concern, reflecting broader societal issues and potential barriers to gender equity. 

This research paper leverages a data set built From six years / 1.55GB of Stack Overflow [survey data](https://survey.stackoverflow.co) to examine the pay 
gap across many variables. 

Our aim is to answer the question, **Is there a significant difference in salary between males and females within the tech industry?**

The dependent variable is *Annual Salary*. We will utilize simple linear regression to first determine if we can reject the null hypothesis. 

The null hypothesis ($H_0$) is: **There is no significant difference in the mean annual salaries between male and females.** 

The alternative hypothesis ($H_1$) is: **There is a significant difference in the mean annual salaries between males and females.** 

Once we determine if we can reject the null hypothesis, we will determine which variables are statistically significant in predicting the response variable (Annual Salary) by creating a multiple linear regression model. 

## Data Overview

```{r, warning = FALSE, message = FALSE, echo = FALSE, show = FALSE, output = FALSE}
source("functions.r")
library(dplyr)
library(psych)
library(gt)
library(stringr)
library(gridExtra)

set.seed(4299)

options(scipen=999)
wide_stack <- get_stack_df(persist = TRUE, load_from_cache = TRUE)

wide_stack <- wide_stack %>%
  mutate(
    python_num = ifelse(python == "yes", 1, 0),
    sql_num = ifelse(sql == "yes",1, 0),
    java_num = ifelse(java == "yes", 1, 0),
    javascript_num = ifelse(javascript == 'yes', 1, 0),
    ruby_num = ifelse(ruby == 'yes',1,0),
    php_num = ifelse(php == 'yes',1,0),
    c_num = ifelse(c == 'yes',1,0),
    swift_num = ifelse(swift == 'yes',1,0),
    scala_num = ifelse(scala == 'yes',1,0),
    r_num = ifelse(r == 'yes',1,0),
    rust_num=ifelse(rust == 'yes',1,0),
    julia_num = ifelse(julia == 'yes',1,0),
    mysql_num = ifelse(mysql == 'yes',1,0),
    microsoftsqlserver_num = ifelse(microsoftsqlserver=='yes',1,0),
    mongodb_num = ifelse(mongodb=='yes',1,0),
    postgresql_num=ifelse(postgresql=='yes',1,0),
    oracle_num = ifelse(oracle=='yes',1,0),
    ibmdb2_num = ifelse(ibmdb2 == 'yes', 1,0),
    redis_num = ifelse(redis=='yes',1,0),
    sqlite_num = ifelse(sqlite=='yes',1,0),
    mariadb_num = ifelse(mariadb=='yes',1,0),
    microsoftazure_num = ifelse(microsoftazure == 'yes',1,0),
    googlecloud_num = ifelse(googlecloud=='yes',1,0),
    ibmcloudorwatson_num = ifelse(ibmcloudorwatson=='yes', 1,0),
    kubernetes_num = ifelse(kubernetes=='yes', 1,0),
    linux_num = ifelse(linux=='yes',1,0),
    windows_num = ifelse(windows == 'yes',1,0),
    aws_num = ifelse(aws == 'yes', 1, 0)
  ) %>%
  filter(Gender %in% c("Female", "Male")) %>%
  filter(Year %in% 2018:2022) %>%
  filter(!is.na(AnnualSalary)) %>% 
  filter(Country == "United States") %>%
  filter(Employment == "Full-Time") %>%
  filter(AnnualSalary < 300000) %>%
  select(-c(Employment, US_State, Sexuality, Ethnicity, OrgSize, YearsCodePro, Age, PlatformWorkedWith, LanguageWorkedWith, DatabaseWorkedWith)) %>%
  mutate(
    Gender = as.factor(Gender),
    EdLevel = as.factor(EdLevel)) %>%
  mutate_at(
    vars(python:aws), 
    as.factor)
```


### Data engineering pipeline

Our [pipeline](https://github.com/tonythor/krijudato/blob/develop/functions.r) does the following: 

1. For each year, download raw survey data file from S3. 
2. Unify columns per year, and then union all years together.
3. Explode wide certain multi-value columns. For example, "PlatformWorkedWith" contains both AWS and Google Cloud.
4. Preprocess certain columns. For example, we aggregate a second grouped_ethnicity column that is only either minority, non-minority, or NA. 
3. Save the CSV file in the root directory so the markdown can load it from cache.

### How we filtered the base data set
We filtered our more than 500K raw records all the way down to 43,655 for this study. Our working dataset includes only those who: 

* Provided their salary

* Work in the United States

* Identify as male or female

* Have full time jobs

* Have an annual salary below 300,000

### A quick glimpse of filtered data

```{r message=FALSE, echo=FALSE}
glimpse(wide_stack)
```

### A high-level respondent overview
The range of respondents includes many types of careers, but is all highly technical in nature. As well, many more men than women responded to this survey, and that most likely implies several different kinds of bias. 

```{r message=FALSE, echo=FALSE}
salByJobType <- wide_stack %>%
  mutate(DevType = str_split_fixed(DevType, ';', 2)[, 1]) %>%
  filter(!is.na(AnnualSalary)) %>%
  group_by(DevType) %>%
  summarise(AvgSalary = mean(AnnualSalary, na.rm = TRUE)) %>%
  arrange(desc(AvgSalary)) %>%
  head(n = 20) %>%
  ggplot(aes(x = reorder(DevType, -AvgSalary), y = AvgSalary)) +
  geom_bar(stat = "identity", position = "dodge") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ggtitle("Top 20 Average Salaries in the U.S. by Job Type") +
  labs(x = "Job Type", y = "Average Salary")

respGender <- wide_stack %>%
  filter(Gender %in% c("Female", "Male")) %>%
  ggplot(aes(x = Year, fill = Gender)) +
  geom_bar(position = "dodge") +
  labs(title = "Respondents by gender over years",
    x = "Year",
    y = "Number of Respondents",
    fill = "Gender") +
  theme_minimal()
```

```{r fig.width=12}
grid.arrange(salByJobType, respGender, ncol = 2)
```

### A non-gender based look at correlation
We converted many independent variables to integers so we could see could look at correlation to the dependent variable.

```{r, warning = FALSE, message = FALSE, echo=FALSE}
numeric_data <- select(wide_stack, Year, AnnualSalary, 
  YearsCodeProAvg, OrgSizeAvg, AgeAvg, python_num,sql_num, java_num, javascript_num, ruby_num, php_num, c_num, swift_num, scala_num, r_num,rust_num, 
  julia_num, mysql_num, microsoftsqlserver_num, mongodb_num, postgresql_num, oracle_num, ibmdb2_num, redis_num, sqlite_num,mariadb_num, 
  microsoftazure_num, googlecloud_num, ibmcloudorwatson_num, kubernetes_num, linux_num, windows_num, aws_num)
cor_matrix <- cor(numeric_data, use = "complete.obs")
cor_with_salary <- cor_matrix["AnnualSalary",]
cor_with_salary <- cor_with_salary[-which(names(cor_with_salary) == "AnnualSalary")]
cor_df <- data.frame(Variable = names(cor_with_salary), Correlation = cor_with_salary)

corr_chart <- ggplot(cor_df, aes(x = Variable, y = Correlation)) +
  geom_bar(stat = "identity") +
  labs(title = "Correlations to Annual Salary", 
       x = "", 
       y = "Correlation coefficient") +
  theme_minimal() +
  ylim(-1, 1) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        plot.title = element_text(face = "bold", hjust = 0.5))
```

```{r fig.width=12}
grid.arrange(corr_chart, ncol = 1)
```


### A quick look at gender discrepency
As a final series of charts, and as we alluded to in the project proposal, the presence of a pay gap was not limited to just one or two variables; rather, it manifested across all of them. Before delving into a more thorough analysis, we decided to incorporate additional descriptive box plots into this presentation. 

```{r message=FALSE, echo=FALSE}
boxplot_base <- wide_stack %>% 
  filter(Gender %in% c("Female", "Male")) %>%
  filter(!is.na(AnnualSalary), !is.na(Gender))

# Filter the base data for each technology
aws_data_yes <- boxplot_base %>% filter(aws == "yes")
linux_data_yes <- boxplot_base %>% filter(linux == "yes")
oracle_data_yes <- boxplot_base %>% filter(oracle == "yes")

miniority <- boxplot_base %>% filter(ethnicity_grouped == "minority")
non_miniority <- boxplot_base %>% filter(ethnicity_grouped == "non-minority")
lgbqt <- boxplot_base %>% filter(sexuality_grouped == "lgbtq")
straight <- boxplot_base %>% filter(sexuality_grouped == "straight")


# Create the AWS plot
aws <- ggplot(aws_data_yes, aes(x = Gender, y = AnnualSalary, fill = Gender)) +
  geom_boxplot() +
  labs(title = "AWS People",
    x = "Gender",
    y = "Annual Salary",
    fill = "Gender") +
  scale_y_continuous(labels = scales::dollar_format()) +
  scale_fill_manual(values = c("Female" = "pink", "Male" = "blue")) +
  theme_minimal()

# Create the Linux plot
linux <- ggplot(linux_data_yes, aes(x = Gender, y = AnnualSalary, fill = Gender)) +
  geom_boxplot() +
  labs(title = "Linux People",
    x = "Gender",
    y = "Annual Salary",
    fill = "Gender") +
  scale_y_continuous(labels = scales::dollar_format()) +
  scale_fill_manual(values = c("Female" = "pink", "Male" = "blue")) +
  theme_minimal()

# Create the Oracle plot
oracle <- ggplot(oracle_data_yes, aes(x = Gender, y = AnnualSalary, fill = Gender)) +
  geom_boxplot() +
  labs(title = "Oracle People",
    x = "Gender",
    y = "Annual Salary",
    fill = "Gender") +
  scale_y_continuous(labels = scales::dollar_format()) +
  scale_fill_manual(values = c("Female" = "pink", "Male" = "blue")) +
  theme_minimal()

# Create the Minority plot
minority_plot <- ggplot(miniority, aes(x = Gender, y = AnnualSalary, fill = Gender)) +
  geom_boxplot() +
  labs(title = "Minorities",
    x = "Gender",
    y = "Annual Salary",
    fill = "Gender") +
  scale_y_continuous(labels = scales::dollar_format()) +
  scale_fill_manual(values = c("Female" = "pink", "Male" = "blue")) +
  theme_minimal()

# Create the Non-Minority plot
non_minority_plot <- ggplot(non_miniority, aes(x = Gender, y = AnnualSalary, fill = Gender)) +
  geom_boxplot() +
  labs(title = "Non-minorities",
    x = "Gender",
    y = "Annual Salary",
    fill = "Gender") +
  scale_y_continuous(labels = scales::dollar_format()) +
  scale_fill_manual(values = c("Female" = "pink", "Male" = "blue")) +
  theme_minimal()

# Create the LGBTQ plot
lgbqt_plot <- ggplot(lgbqt, aes(x = Gender, y = AnnualSalary, fill = Gender)) +
  geom_boxplot() +
  labs(title = "LGBTQ",
    x = "Gender",
    y = "Annual Salary",
    fill = "Gender") +
  scale_y_continuous(labels = scales::dollar_format()) +
  scale_fill_manual(values = c("Female" = "pink", "Male" = "blue")) +
  theme_minimal()

# Create the Straight plot
straight_plot <- ggplot(straight, aes(x = Gender, y = AnnualSalary, fill = Gender)) +
  geom_boxplot() +
  labs(title = "Straight",
    x = "Gender",
    y = "Annual Salary",
    fill = "Gender") +
  scale_y_continuous(labels = scales::dollar_format()) +
  scale_fill_manual(values = c("Female" = "pink", "Male" = "blue")) +
  theme_minimal()
```

```{r fig.width=12}
grid.arrange(aws, linux, oracle, ncol = 3)
grid.arrange(minority_plot, non_minority_plot, lgbqt_plot , straight_plot, ncol = 4)
```



## Data analysis

### Simple Linear Regression

To determine if we can reject the null hypothesis, we will begin with constructing a simple linear regression model with annual salary as the dependent variable and gender as the independent variable.

```{r}
wide_stack$Gender <- relevel(wide_stack$Gender, ref = "Male")
m_salary_gender <- lm(AnnualSalary ~ Gender, data = wide_stack)
summary(m_salary_gender)
```

From the above, we see that the p-value is 0.000000000000022%. Using a 95% confidence interval, having a p-value of less than 5% means that we can reject the null hypothesis. We can see that the adjusted $R^2$ value is very small (0.6364%), which means that the model does not explain much of the variability of the dependent variable. The coefficient -14,112.9 means that, on average and after accounting for other factors in the model, being female is associated with a decrease of $14,112.9 in annual salary compared to males.

---

### Multiple Linear Regression

Now that we have determined that we can reject the null hypothesis, we will construct a multiple linear regression model which includes all of the variables in our study to determine which are good predictors of the variance in annual salary. The inclusion of interaction terms, represented as "Gender * Variable," allows us to examine the impact of each variable in relation to gender.

```{r}
wide_stack$EdLevel<- relevel(wide_stack$EdLevel, ref = "Something Else")

m_salary <- lm(AnnualSalary ~ Gender + Gender:AgeAvg + Gender:ethnicity_grouped + Gender:sexuality_grouped +
   Gender:EdLevel + Gender:OrgSizeAvg +
   Gender:YearsCodeProAvg + Gender:Year + Gender:python + Gender:r + Gender:scala + Gender:julia +
   Gender:microsoftazure + Gender:aws + Gender:mariadb + Gender:mongodb +
   Gender:linux + Gender:windows + Gender:mysql + Gender:oracle + Gender:ibmdb2 + 
   Gender:c + Gender:googlecloud + Gender:ibmcloudorwatson + Gender:java + 
   Gender:javascript + Gender:kubernetes + + Gender:microsoftsqlserver + Gender:php +
   Gender:postgresql + Gender:redis + Gender:ruby + Gender:rust + Gender:sqlite +
   Gender:swift, data = wide_stack)

summary(m_salary)
```

The $R^2$ value is 30.25%, which means that approximately 30.25% of the variability in the dependent variable (AnnualSalary) can be explained by the independent variables included in the model. The remaining 69.75% of the variability is unaccounted for by the model. The p-value is 0.000000000000022%,  which suggests that there is strong evidence that at least one of the predictors in the model has a non-zero effect, and the overall model is statistically significant. Hence, we can reject the null hypothesis (There is no significant difference in the mean annual salaries between male and females).

## Assuring compliance with model conditions

Having examined the four essential conditions for multiple linear regression, including linearity, normality, constant variability, and independence of residuals, we conclude that the utilization of multiple linear regression is deemed valid.

### Linearity

We can assume linearity, as there is no apparent trend observed in the distribution.

```{r, message = FALSE, warning = FALSE}
ggplot(data = m_salary, aes(x = .fitted, y = .resid)) +
  geom_point()
```

### Normality
We can assume normality as the line mostly falls on the normal line. 

```{r, warning = FALSE, message = FALSE}
ggplot(data = m_salary, aes(sample = .resid)) +
  stat_qq() +
  stat_qq_line()
```

### Constant Variability

Points are scattered with no apparent pattern around 0, indicating that we can assume constant variability.

```{r}
ggplot(data = m_salary, aes(x = .fitted, y = .resid)) +
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed")
```


### Independence

Given that each case represents an individual response, we can reasonably assume independence. Even though the same individuals may submit survey responses over multiple years, we can expect their salaries to change within a year due to factors such as bonuses, promotions, and annual salary increases.

## Conclusion

### Findings
Having analyzing gigabytes of data spanning 557 raw variables, which were expanded to over 750 variables, the following conclusions have become evident:

1. The linear model that we've developed only accounts for 30.25% of the variability of the dependent variable
2. There is another variable, or variables, not previously considered that plays a pivotal role in describing the wage gap within the technology industry

In conclusion, there is an obvious gap in salaries between males and females shown in our regression model. 

For example, the coefficients for GenderFemale:awsyes and GenderMale:awsyes represent the estimated change in Annual Salary associated with a one-unit change in the variable "awsyes", while holding all other variables constant. The coefficients for GenderFemale:awsyes and GenderMale:awsyes are 9,247 and 10,137, respectively. Both coefficients are positive, which means that, on average, both men and women who use AWS tend to have higher salaries compared to those who do not use AWS. However, the magnitude of the coefficient for GenderMale:awsyes is higher than that for GenderFemale:aws yes. This suggests that, on average, men who use AWS have a higher estimated increase in Annual salary compared to women who use AWS.

This trend is shown across several variables in our model. In conjunction with the results from our simple linear regression model, we can conclude that our model supports our alternative hypothesis that there is a significant difference in the mean annual salaries between males and females. Importantly, this difference in mean annual salaries is not confined to a single variable but holds true across multiple variables considered in the study.

### Recommendations
We anticipate that delving deeper into the factors examined in this study may yield inconclusive results. However, our main suggestion is to persist with this research, seek funding, or take any necessary measures to ensure its continuation. We have identified three potential directions to pursue next. Nonetheless, we are confident that exploring these avenues would likely uncover influential variables contributing to the pay gap.

1. **Study salaries by gender and location**  Our study lacks consideration for state and county variables. It's plausible that the wage gap is specific to small towns versus big cities, or possibly confined to a particular location such as Portland, Oregon. The current data does not incorporate this geographical dimension.

2. **Study company salaries by gender** Certainly, not every company, regardless of its size, faces substantial issues with pay disparities. However, specific companies are notorious for having a significantly massive wage gap. Examining the salaries of individual companies and discerning the wage gap within each could be a highly impactful analysis.

3. **Study entire industries by gender and salary** Software developers and project managers performing identical tasks are found in both non-profit and media sectors. It's possible that the industry itself plays a pivotal role in influencing the wage gap.



