---
title: "Flood Risk in NYC: A Product for Insurance Professionals"
author: "Team Krijudato (Kristin L, Julia F, David G, Tony F)"
date: "December 10, 2023"
format:
  html:
    theme: cosmo
    toc: true
    number_sections: true
---

## Introduction

do a bunch of stuff, two data sets, blah

## load stuff

```{r, warning = FALSE, message = FALSE, echo = FALSE, show = FALSE, output = FALSE}
library(tidyverse)
library(scales)
library(dplyr)
library(gt)
library(leaflet)
library(htmltools)
joined_pluto_path <- "./nogit_cache/_joined_pluto.csv"
if (!file.exists(joined_pluto_path)) {
  source('./nycfun.r')
}

## load the cache
joined_pluto <- read_csv(joined_pluto_path) %>% 
  mutate(pfirm15_flag = coalesce(pfirm15_flag, 0))

```

```{r fig.height=6}
joined_pluto_sorted <- joined_pluto %>%
  arrange(desc(property_value))

ggplot(joined_pluto_sorted, aes(x = reorder(location_name, property_value), y = property_value)) +
  geom_bar(stat = "identity") +
  scale_y_continuous(labels = dollar_format(prefix = "$", suffix = "", big.mark = ",", decimal.mark = ".", accuracy = 1)) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  labs(x = "Location Name", y = "Property Value", title = "Property Value by Location")
```

## Some Maps

## The Product Output

According to [Swiss Re](https://www.swissre.com/risk-knowledge/risk-perspectives-blog/challenge-of-understanding-secondary-perils.html), flooding is classified as a secondary peril, posing greater modeling challenges compared to hurricanes and earthquakes. The unpredictability of flood occurrences, both in terms of timing and location, adds to the complexity. The limited adoption of flood catastrophe models by insurance professionals can be attributed to these difficulties. Current flood catastrophe models do not have a high degree of confidence, as U.S. flood models are still in their infancy. 

The product we've created proves valuable for insurance professionals by offering insights into the locations outlined in a schedule of values. While it doesn't fall under the category of a catastrophe model, it furnishes valuable information. This information aids insurance professionals, including underwriters and actuaries, in making well-informed decisions regarding the provision of flood insurance for individuals, groups, companies, or government entities. 

In our demonstration, we are using the schedule of values for the City University of New York (CUNY). 

### Map of Locations

```{r, warning = FALSE, message = FALSE, show = FALSE, echo = FALSE}
locations <- data.frame(
  location_name = joined_pluto_sorted$location_name,
  latitude = joined_pluto_sorted$latitude,
  longitude = joined_pluto_sorted$longitude,
  borough = joined_pluto_sorted$borough.y,
  assesstot = joined_pluto_sorted$assesstot,
  pfirm15_flag = joined_pluto_sorted$pfirm15_flag,
  bldgarea = joined_pluto_sorted$bldgarea,
  lotarea = joined_pluto_sorted$lotarea,
  unitstotal = joined_pluto_sorted$unitstotal,
  numbldgs = joined_pluto_sorted$numbldgs,
  numfloors = joined_pluto_sorted$numfloors,
  yearbuilt = joined_pluto_sorted$yearbuilt,
  out_condo_flag = joined_pluto_sorted$out_condo_flag
)

popup_content <- function(location_name, borough, assesstot, bldgarea, lotarea, unitstotal, numbldgs, numfloors, yearbuilt, pfirm15_flag, out_condo_flag) {
  flood_zone <- ifelse(pfirm15_flag == 0, "Not in Flood Zone", "1% Annual Flood Plain")
  paste(
    "<b>Location:</b> ", location_name, "<br>",
    "<b>Borough:</b> ", borough, "<br>",
    "<b>Property Value:</b> ", scales::dollar(assesstot), "<br>",
    "<b>Building Area:</b> ", scales::comma(bldgarea), "<br>",
    "<b>Lot Area:</b> ", scales::comma(lotarea), "<br>",
    "<b>Total Units:</b> ", unitstotal, "<br>",
    "<b>Number of Buildings:</b> ", numbldgs, "<br>",
    "<b>Number of Floors:</b> ", numfloors, "<br>",
    "<b>Year Built:</b> ", yearbuilt, "<br>",
    "<b>Flood Zone:</b> ", flood_zone, "<br>",
    "<b>Type:</b> ", out_condo_flag
  )
}

loc_map <- leaflet(locations) %>%
  addTiles() %>%
  addMarkers(
    ~longitude, ~latitude,
    popup = popup_content(
      joined_pluto_sorted$location_name,
      joined_pluto_sorted$borough.y,
      joined_pluto_sorted$assesstot,
      joined_pluto_sorted$bldgarea,
      joined_pluto_sorted$lotarea,
      joined_pluto_sorted$unitstotal,
      joined_pluto_sorted$numbldgs,
      joined_pluto_sorted$numfloors,
      joined_pluto_sorted$yearbuilt,
      joined_pluto_sorted$pfirm15_flag,
      joined_pluto_sorted$out_condo_flag
    ),
    labelOptions = labelOptions(noHide = TRUE, textOnly = TRUE)  
  )

loc_map
```

### Property Value by Flood Zone

```{r, message = FALSE, warning = FALSE, echo = FALSE, show = FALSE, fig.align="center"}
# property value per flood zone
joined_pluto_sorted %>%
  group_by(pfirm15_flag) %>%
  summarize(total_assesstot = sum(assesstot, na.rm = TRUE)) %>%
  ggplot() +
  geom_bar(aes(x = factor(pfirm15_flag), y = total_assesstot, fill = factor(pfirm15_flag)), stat = "identity", position = "dodge") +
  geom_text(aes(x = factor(pfirm15_flag), y = total_assesstot, label = comma_format()(total_assesstot)),
            position = position_dodge(width = 0.9),
            vjust = -0.5, size = 3) +
  labs(title = "Total Values by Flood Risk Type",
       x = "",
       y = "Total Values") +
  scale_x_discrete(labels = c("0" = "No Flood Risk", "1" = "1% Annual Chance Flood Plain")) +
  scale_y_continuous(labels = comma_format(), breaks = seq(0, 2000000000, by = 250000000)) +
  scale_fill_manual(values = c("0" = "lightgreen", "1" = "indianred"), 
                    name = "Flood Zone Risk", labels = c("No Flood Risk", "1% Annual Chance Flood Plain")) +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold", hjust = 0.5),
        legend.title = element_text(face = "bold", hjust = 0.5))
```

### Risk Scoring

Risk scores can prove to be a useful tool for underwriters to determine the complexity of the flood risk for the schedule. We've developed a risk scoring system that is assigned as follows:

- Properties *not* in a flood zone and *without* a basement receive a risk score of **0**.
- Properties *not* in a flood zone but *with* a basement are assigned a risk score of **1**.
- Properties *in* a flood zone and *without* a basement are given a risk score of **2**.
- Properties *in* a flood zone and *with* a basement receive a risk score of **3**.

```{r, message = FALSE, warning = FALSE, echo = FALSE, show = FALSE, fig.align="center"}
# provide risk score
# not in flood zone and doesn't have basement = 0
# not in flood zone but has basement = 1
# in a flood zone = 2
# in a flood zone & has basement  = 3
risky_prop_flood <- joined_pluto_sorted %>%
  mutate(
    riskscore = case_when(
      pfirm15_flag == 1 & (!is.na(bsmtcode) | bsmtcode != 0) ~ 3,
      pfirm15_flag == 1 & (is.na(bsmtcode) | bsmtcode == 0) ~ 2,
      pfirm15_flag == 0 & (!is.na(bsmtcode) | bsmtcode != 0) ~ 1,
      pfirm15_flag == 0 & (is.na(bsmtcode) | bsmtcode == 0) ~ 0,
      TRUE ~ NA_real_
    )
  ) %>%
  select(location_name, street_number, city, state, zip, borough.x, riskscore, assesstot)


ggplot(data = risky_prop_flood) +
  geom_bar(aes(x = riskscore, fill = as.factor(riskscore))) +
  labs(
    title = "Distribution of Risk Scores",
    x = "Risk Score",
    y = "Count"
  ) +
  scale_y_continuous(breaks = seq(0, max(table(risky_prop_flood$riskscore)), by = 5)) +
  scale_fill_manual(values = c("darkolivegreen2", "gold1", "sienna1", "firebrick"),
                    breaks = c(0, 1, 2, 3),
                    labels = c("0", "1", "2", "3"),
                    name = "Risk Score") +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5)
  )

```

### Summary of Values by Risk Score

```{r, warning = FALSE, message = FALSE, echo = FALSE, show = FALSE, fig.align="center"}
summary_data <- risky_prop_flood %>%
  group_by(riskscore) %>%
  summarize(total_assesstot = sum(assesstot, na.rm = TRUE)) %>%
  rename("New Name for Total Assesstot" = total_assesstot,
         "Risk Score" = riskscore)

summary_data %>%
  gt() %>%
  cols_label("Risk Score" = md("**Risk Score**"),
             "New Name for Total Assesstot" = md("**Values**")) %>%
  tab_header(
    title = md("**Summary of Values by Risk Score**")
  ) %>%
  fmt_currency(columns = c("New Name for Total Assesstot"), currency = "USD")
```
## Map of Flood Zones in NYC

```{r flood-map, message = FALSE, fig.align = "center"}
url <- "https://tonyfraser-data.s3.amazonaws.com/nyc-addresses/nyc_pluto_23v3_csv/pluto_23v3.csv"
pluto_path <- "_pluto_23v3.csv"
if (!file.exists(pluto_path)) {
  download.file(url, pluto_path, mode = "wb")
}

pluto <- read_csv(pluto_path)
dfbrklyn<-filter(pluto,borocode=='3', pfirm15_flag=='1')
dfqueens<-filter(pluto,borocode=='4', pfirm15_flag=='1')
dfbronx<-filter(pluto,borocode=='2', pfirm15_flag=='1')
dfmanhattan<-filter(pluto,borocode=='1', pfirm15_flag=='1')
dfstaten<-filter(pluto,borocode=='5', pfirm15_flag=='1')

df_sample_Brooklyn<-sample_n(dfbrklyn,2192)
df_sample_Queens<-sample_n(dfqueens,2155)
df_sample_Bronx<-sample_n(dfbronx,1399)
df_sample_manhattan<-sample_n(dfmanhattan,1283)
df_sample_staten<-sample_n(dfstaten,2040)

df_sample_Brooklyn_flood<- df_sample_Brooklyn %>%
  mutate(Brooklyn_Data= paste("Building Area: ", bldgarea,
  "Lot Area: ", lotarea,
  "Total Number of Units: ", unitstotal,
  "Number of Buildings: ", numbldgs,
  "Number of Floors: ", numfloors,
  "Assessed Total: ", assesstot,
  "Year Built: ", yearbuilt,
  "Location: ",address))%>%mutate(County=paste("Kings County"))

df_sample_Queens_flood<- df_sample_Queens %>%
  mutate(Queens_Data= paste("Building Area: ", bldgarea,
  "Lot Area: ", lotarea,
  "Total Number of Units: ", unitstotal,
  "Number of Buildings: ", numbldgs,
  "Number of Floors: ", numfloors,
  "Assessed Total: ", assesstot,
  "Year Built: ", yearbuilt,
  "Location: ",address))%>%mutate(County=paste("Queens County"))

df_sample_Bronx_flood<- df_sample_Bronx %>%
  mutate(Bronx_Data= paste("Building Area: ", bldgarea,
  "Lot Area: ", lotarea,
  "Total Number of Units: ", unitstotal,
  "Number of Buildings: ", numbldgs,
  "Number of Floors: ", numfloors,
  "Assessed Total: ", assesstot,
  "Year Built: ", yearbuilt,
  "Location: ",address))%>%mutate(County=paste("Bronx County"))

df_sample_manhattan_flood<- df_sample_manhattan %>%
  mutate(Manhattan_Data= paste("Building Area: ", bldgarea,
  "Lot Area: ", lotarea,
  "Total Number of Units: ", unitstotal,
  "Number of Buildings: ", numbldgs,
  "Number of Floors: ", numfloors,
  "Assessed Total: ", assesstot,
  "Year Built: ", yearbuilt,
  "Location: ",address))%>%mutate(County=paste("New York County"))

df_sample_staten_flood<- df_sample_staten %>%
  mutate(Staten_Island_Data= paste("Building Area: ", bldgarea,
  "Lot Area: ", lotarea,
  "Total Number of Units: ", unitstotal,
  "Number of Buildings: ", numbldgs,
  "Number of Floors: ", numfloors,
  "Assessed Total: ", assesstot,
  "Year Built: ", yearbuilt,
  "Location: ",address))%>%mutate(County=paste("Richmond County"))

sample_colleges <- joined_pluto %>%
  mutate(colleges_data = paste("Building Area: ", bldgarea,
  "Lot Area: ", lotarea,
  "Total Number of Units: ", unitstotal,
  "Number of Buildings: ", numbldgs,
  "Number of Floors: ", numfloors,
  "Assessed Total: ", assesstot,
  "Year Built: ", yearbuilt,
  "Location: ",address))

m_df_sample_staten_flood<-mapview(df_sample_staten_flood,crs = 4269,  xcol = "longitude", ycol = "latitude",zcol="County", legend = TRUE, popup="Staten_Island_Data", grid = FALSE, layer.name = "Richmond County", col.regions = "red")

m_df_sample_manhattan_flood<-mapview(df_sample_manhattan_flood,crs = 4269,  xcol = "longitude", ycol = "latitude", zcol="County",popup="Manhattan_Data",legend = TRUE,  grid = FALSE, layer.name = "New York County", col.regions = "green")

m_df_sample_Bronx_flood<-mapview(df_sample_Bronx_flood,crs = 4269,  xcol = "longitude", ycol = "latitude",zcol="County", legend = TRUE, popup="Bronx_Data",grid = FALSE, layer.name = "Bronx County", col.regions = "blue" )

m_df_sample_Queens_flood<-mapview(df_sample_Queens_flood,crs = 4269,  xcol = "longitude", ycol = "latitude",zcol="County", legend = TRUE, popup="Queens_Data",grid = FALSE, layer.name = "Queens County", col.regions = "orange" )

m_df_sample_Brooklyn_flood<-mapview(df_sample_Brooklyn_flood,crs = 4269,  xcol = "longitude", ycol = "latitude",zcol="County", legend = TRUE, popup="Brooklyn_Data", grid = FALSE, layer.name = "Kings County", col.regions = "yellow")

colleges <- mapview(sample_colleges, crs = 4269, xcol = "longitude", ycol= "latitude", zcol = "type", legend = TRUE, popup = "colleges_data", grid = FALSE, layer.name = "Colleges in NYC", col.regions = c("black", "gray", "white"))

library(readr)
polygonData <- read_csv("polygonData.csv")
# The data in the file polygonData.csv was uploaded from this website: https://data.cityofnewyork.us/Environment/Floodplain/g62a-qs52
# Then, click Export and download the CSV file under the section "Download a non-geospatial file type"
# Save it to polygonData.csv and then use it for this code

polygonData <- polygonData |> filter(Shape_Area >= 1)

geometry1 <- strsplit(polygonData$the_geom, split = ",")
geometry2a <- str_split(geometry1, ",")
geometry3 <- lapply(geometry2a, as.data.frame)

df2 <- c()

for (i in 1:nrow(polygonData))
{
  df1 <- data.frame(XY = geometry3[[i]], num = rep(i, each = nrow(geometry3[[i]])))
  df2 <- rbind(df1, df2)
}

df2$`X..i..` <- gsub('c\\(\\"MULTIPOLYGON \\(\\(\\(', "", df2$`X..i..`)
df2$X..i.. <- gsub('\\)\\)\\)\\"\\)', '', df2$X..i..)
df2$X..i.. <- gsub('\\"', '', df2$X..i..)
df2$X..i.. <- gsub('\\\n', '', df2$X..i..)
df2$X..i.. <- gsub(' 4', ' 44', df2$X..i..)
df2$X..i.. <- gsub('\\(', '', df2$X..i..)
df2$X..i.. <- gsub('\\)', '', df2$X..i..)

df2[c('long', 'lat')] <- str_split_fixed(df2$X..i.., ' 4', 2)
df2 <- df2[,-1]

projcrs <- "+proj=longlat +datum=WGS84 +no_defs +ellps=WGS84 +towgs84=0,0,0"
myCoords <- st_as_sf(x = df2,
                     coords = c("long", "lat"),
                     crs = projcrs)

polys <- st_sf(
  aggregate(
    myCoords$geometry,
    list(testing$num),
    function(g){
       st_cast(st_combine(g),"POLYGON")
    }
   ))

polygons1 <- mapview(polys)

m_df_sample_Brooklyn_flood+m_df_sample_Queens_flood+m_df_sample_Bronx_flood+m_df_sample_manhattan_flood+m_df_sample_staten_flood + colleges + polygons1
```

## Conclusion

## Future Improvements

Improving this product in the future may entail incorporating data at the national level and potentially expanding it globally. However, challenges arise when applying it nationally and globally due to varying levels of detail in flood information from different countries and regions. 

Another improvement would be to add another data source that provides building construction. The type of construction can significantly impact the extent of losses in the event of a flood. Assessing the materials and design of structures would help insurance professionals estimate potential damage and calculate appropriate coverage.
